---
title: "COVID-19 Data Analysis Report"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# repos="https://utstat.toronto.edu/cran/"
library(tidyverse)
library(stringr)
library(lubridate)
library(maps)
```

# Project introduction

**What factors can be related to the level of observed infection / recovery / death by COVID-19 at a given time and given region?**
**How has COVID-19 impacted society in terms of economic changes and lifestyle trends?**

In the following, besides the **4** online data sets on COVID-19, there are **21** data sets concerning a number of potential factors of interest, such as *educational*, *political*, *economical*, *employment*, *health*, *demographic*, *infrastructure*, *self perception* factors. A number of COVID-19 related data sets are included, which by the collective work of many organizations, are updated real-time. The *WorldRegions.csv* data from World Regions Classification list are included.

Also included are the data *WorldHappinessReport2021-Score.csv* from the World Happiness Report `2021`, which concerns the years `2018-2020`. It is computed based on the answers of people to the following question: “Please imagine a ladder, with steps numbered from 0 at the bottom to 10 at the top. The top of the ladder represents the best possible life for you and the bottom of the ladder represents the worst possible life for you. On which step of the ladder would you say you personally feel you stand at this time?” ([Statistical Appendix 1 for Chapter 2](https://happiness-report.s3.amazonaws.com/2021/Appendix1WHR2021C2.pdf) of [World Health Report 2021](https://worldhappiness.report/ed/2021/)) Thus, the score can be seen as giving one interpretation of happiness.

## The Map

The map below shows the `new cases` on Oct. 31, 2021 obtained from [Our world in data](https://covid.ourworldindata.org/data/owid-covid-data.csv). 

```{r include=FALSE}
COVID_cases <- read_csv("COVID-2021-10-31.csv",
                        col_types = cols(
                          location = col_character(),
                          new_cases = col_double(),
                          total_cases = col_double()
                        ))
```

```{r echo=FALSE, warning=FALSE}
world <- map_data("world")
iu <- COVID_cases %>% rename (region = location)
 # to match world map data, see comments below
iu$region[42] <- "Republic of Congo"
iu$region[44] <- "Ivory Coast"
iu$region[48] <- "Czech Republic"
iu$region[49] <- "Democratic Republic of the Congo"
iu$region[64] <- "Faroe Islands"
iu$region[128] <- "Micronesia"
iu$region[194] <- "Timor"
iu$region[203] <- "UK"
iu$region[204] <- "USA"
iu <- semi_join(iu, world, by = "region") #only keep countries according to world map data
gg <- ggplot()
gg <- gg + geom_map(
  data = world,
  map = world,
  aes(x = long, y = lat, map_id = region),
  fill = "#ffffff",
  color = "#ffffff",
  size = 0.20
  )
  
  gg <- gg + geom_map(
  data = iu,
  map = world,
  aes(fill = new_cases, map_id = region),
  color = "#ffffff",
  size = 0.15
  )
  
  gg <- gg + scale_fill_continuous(low = 'thistle2', high = 'darkblue',
  guide = 'colorbar')
  gg
```

# The Setup:

Data analysis date range: Between January 2020 and December 2021

# The Data sets

There are a total of **27** data sets, **4** of which are online real-time data sets that are regularly updated.

- CIA World FactBook: from which we obtained a majority of the data sets as `.csv` files
- World Bank Data: from which we obtained the data sets on Freshwater Resources, Internet Usage, Health Expenditure, Urbanization, Labor Force and some Course of Death data
- Our World in Data: from which we obtain the online data sets on COVID-19 cases, vaccinations and testing
- [Economist Intelligence Unit](http://www.eiu.com/home.aspx): which developed the [democracy index](https://www.eiu.com/topic/democracy-index). The version we use is from the [Wikipedia page](https://en.wikipedia.org/wiki/Democracy_Index).
- University of Oxford: from which we obtain the online data set on Government Response
- United Nations: from which we obtained the data set on Population Distribution by Age and Gender
- Wikipedia: from which we obtained the World Regions Classification data set, aside from the democracy index mentioned above

### Real-time COVID-19 data sets:

These data are regularly updated, and they contain all the historical data, which include the periods that we are interested in. 

- `covid_complete`: [COVID-19 cases (Our world in data)](https://covid.ourworldindata.org/data/owid-covid-data.csv)
```{r results='hide', warning=FALSE, include=FALSE}
# need to remove include=FALSE above when send out as project file
covid_cases <- read_csv("https://covid.ourworldindata.org/data/owid-covid-data.csv",
                        col_types = cols(
                          .default = col_double(),
                          date = col_date(format =""),
                          location = col_character()
                        ))
covid_cases %>% head()
```

- `covid_vaccinations`: [COVID-19 vaccinations (Our world in data)](https://raw.githubusercontent.com/owid/covid-19-data/master/public/data/vaccinations/vaccinations.csv)
  Some portion of this data is included in the `covid_complete` data above.

```{r results='hide', warning=FALSE, message=FALSE, include=FALSE}
# need to remove include=FALSE above when send out as project file
covid_vaccs <- read_csv("https://raw.githubusercontent.com/owid/covid-19-data/master/public/data/vaccinations/vaccinations.csv")
covid_vaccs %>% head()
```

- `covid_tests`: [COVID-19 testing (Our world in data)](https://covid.ourworldindata.org/data/testing/covid-testing-all-observations.csv)
  Some portion of this data is included in `covid_complete` in the above.

    Hasell, J., Mathieu, E., Beltekian, D. et al. A cross-country database of COVID-19 testing. Sci Data 7, 345 (2020). https://doi.org/10.1038/s41597-020-00688-8

```{r results='hide', warning=FALSE, message=FALSE, include=FALSE}
# need to remove include=FALSE above when send out as project file
covid_tests <- read_csv("https://covid.ourworldindata.org/data/testing/covid-testing-all-observations.csv")
covid_tests %>% head()
```

- `covid_response`: [COVID-19 government responses](https://www.bsg.ox.ac.uk/research/research-projects/coronavirus-government-response-tracker#data)

    Thomas Hale, Noam Angrist, Rafael Goldszmidt, Beatriz Kira, Anna Petherick, Toby Phillips, Samuel Webster, Emily Cameron-Blake, Laura Hallas, Saptarshi Majumdar, and Helen Tatlow. (2021). “A global panel database of pandemic policies (Oxford COVID-19 Government Response Tracker).” Nature Human Behaviour. https://doi.org/10.1038/s41562-021-01079-8

```{r include=FALSE}
#results='hide', warning=FALSE, include=FALSE}
# need to remove include=FALSE above when send out as project file
covid_response <- read_csv("https://raw.githubusercontent.com/OxCGRT/covid-policy-tracker/master/data/OxCGRT_latest.csv",
                           col_types = cols(
                             .default = col_character(),
                             Date = col_date(format = "%Y%m%d"))
                           ) %>%
    mutate(
      across(starts_with("Confirmed"), as.integer),
      across(ends_with("Flag"), as.logical), 
      across(ends_with("Index"), as.double), 
      across(ends_with("ForDisplay"), as.double),
      )
covid_response %>% head()
```

#### Datasets contained in `.csv` files

- `democracyindex`: [democracy index](https://www.eiu.com/topic/democracy-index) developed by the [Economist Intelligence Unit](http://www.eiu.com/home.aspx), which is contained in the table from the [Wikipedia page](https://en.wikipedia.org/wiki/Democracy_Index).
```{r include=FALSE}
democracyindex <- read_tsv("DEMOCRACYINDEX.csv")
democracyindex %>% head()
```

- `regionclassification`: [World Regions Classification](https://meta.wikimedia.org/wiki/List_of_countries_by_regional_classification)
```{r include=FALSE}
regionclassification <- read_tsv("WorldRegions.csv")
regionclassification %>% head()
```

- `happinessscore`: [World happiness report 2021, happiness score](https://worldhappiness.report/ed/2021/)
```{r include=FALSE}
happinessscore <- read_tsv("WorldHappinessReport2021-Score.csv")
happinessscore %>% head()
```

- `area`: [Area of the regions](https://www.cia.gov/the-world-factbook/field/area/country-comparison) in square kilometers
```{r include=FALSE}
area <- read_tsv("AREA.csv")
area %>% head()
```

- `population`: [Population in the region](https://www.cia.gov/the-world-factbook/field/population/country-comparison)
```{r include=FALSE}
population <- read_tsv("POPULATION.csv")
population %>% head()
```

- `populationdistribution`: [Population distribution](https://population.un.org/wpp/Download/Files/1_Indicators%20(Standard)/CSV_FILES/WPP2019_PopulationBySingleAgeSex_1950-2019.csv)
The numeric values for the Pop columns are population in thousands, as of `2019`.
```{r include=FALSE}
populationdistribution <- read_tsv("POPULATIONDISTRIBUTION.csv")
populationdistribution %>% head()
```

- `lifeexpect`: [Life expectance at birth in the region](https://www.cia.gov/the-world-factbook/field/life-expectancy-at-birth)
The most up-to-date data is not as detailed, containing both male and female separately. The data is from year `2021`.
```{r include=FALSE}
lifeexpect <- read_tsv("LIFEEXPECTANCYATBIRTH.csv")
lifeexpect %>% head()
```

- `birthrate`: [Birthrate in the regions](https://www.cia.gov/the-world-factbook/field/birth-rate/country-comparison) The value is the birth rate per `1000` population.
```{r include=FALSE}
birthrate <- read_tsv("BIRTHRATE.csv")
birthrate %>% head()
```

- `deathrate`: [Deathrate in the region](https://www.cia.gov/the-world-factbook/field/death-rate/country-comparison) The value is the death rate per `1000` population.
```{r include=FALSE}
deathrate <- read_tsv("DEATHRATE.csv")
deathrate %>% head()
```

- `causeofdeath`: [Cause of death, by communicable diseases and maternal, prenatal and nutrition conditions (as percentage of total deaths)](https://data.worldbank.org/indicator/SH.DTH.COMM.ZS) The values are the percentage of the total deaths in the region that cause by the specific reasons.
```{r include=FALSE}
causeofdeath <- read_csv("CAUSEOFDEATH.csv")
causeofdeath %>% head()
```

- `laborforce`: [Labor force data in the region](https://data.worldbank.org/indicator/SL.TLF.TOTL.IN) the up-to-date data we use.
[CIA](https://www.cia.gov/the-world-factbook/field/labor-force/country-comparison) has a slightly outdated data set.
```{r include=FALSE}
laborforce <- read_csv("LABORFORCE.csv",
                       col_types = cols(
                         .default = col_double(),
                         `Country Name` = col_character(),
                         `Country Code` = col_character()
                       ))
laborforce %>% head()
```

- `unemployment`: [Unemployment in the region](https://www.cia.gov/the-world-factbook/field/unemployment-rate/country-comparison) Pretty outdated, while it is better than nothing.
```{r include=FALSE}
unemployment <- read_tsv("UNEMP.csv")
unemployment %>% head()
unemployment
covid_cases
```

- `unemp_youth`: [Unemployment of youth in the region, ages 15-24](https://www.cia.gov/the-world-factbook/field/unemployment-youth-ages-15-24/country-comparison) The value is the percentage of youth between 15 and 24 that are unemployed.
```{r include=FALSE}
unemp_youth <- read_tsv("UNEMPYOUTH.csv")
unemp_youth %>% head()
```

- `urbanization`: [Degree of urbanization in the region](https://data.worldbank.org/indicator/SP.URB.TOTL.IN.ZS) The values are percentage of the total population in urban areas.
```{r include=FALSE}
urbanization <- read_tsv("URBANIZATION.csv")
urbanization %>% head()
```

- `healthexp`: [Health expenditure in the region](https://data.worldbank.org/indicator/SH.XPD.CHEX.PC.CD) The values are health expenditure per capital, in terms of current US dollar values.
```{r include=FALSE}
healthexp <- read_tsv("HEALTHEXP.csv")
healthexp %>% head()
```

- `educationexp`: [Education expenditure in the region](https://www.cia.gov/the-world-factbook/field/education-expenditures/country-comparison) The values in the table are in `% of GDP` of the region.
```{r include=FALSE}
educationexp <- read_tsv("EDUEXP.csv")
educationexp %>% head()
```

- `gdppp`: [GDP per capital in the region](https://www.cia.gov/the-world-factbook/field/real-gdp-per-capita/country-comparison)
```{r include=FALSE}
gdppp <- read_tsv("GDPPP.csv")
gdppp %>% head()
```

- `publicdebt`: [Public debt in the region](https://www.cia.gov/the-world-factbook/field/public-debt/country-comparison) The value is in `% of GDP` of the region.
```{r include=FALSE}
publicdebt <- read_tsv("PUBLICDEBT.csv")
publicdebt %>% head()
```

- `gdpcomp_origin`: [GDP composition by sector of origin in the region](https://www.cia.gov/the-world-factbook/field/gdp-composition-by-sector-of-origin/)
```{r include=FALSE}
gdpcomp_origin <- read_tsv("GDPCOMPOSITION-ORIGIN.csv")
gdpcomp_origin %>% head()
```

- `gdpcomp_enduse`: [GPD composition by sector of end use in the region](https://www.cia.gov/the-world-factbook/field/gdp-composition-by-end-use/)
```{r include=FALSE}
gdpcomp_enduse <- read_tsv("GDPCOMPOSITION-ENDUSE.csv")
gdpcomp_enduse %>% head()
```

- `gini` [GINI index in the region](https://www.cia.gov/the-world-factbook/field/gini-index-coefficient-distribution-of-family-income/country-comparison)
```{r include=FALSE}
gini <- read_tsv("GINI.csv")
gini %>% head()
```

- `freshwaterresource` [Renewable internal freshwater resources per capita](https://data.worldbank.org/indicator/ER.H2O.INTR.PC) The values are renewable internal freshwater resources per capita, in cubic meters. It refers to internal river flow and groundwater from rainfall in the region.
```{r include=FALSE}
freshwaterresource <- read_tsv("FRESHWATERRESOURCE.csv")
freshwaterresource %>% head()
```

- `internetuser`: [Individuals using the Internet (as percentage of population)](https://data.worldbank.org/indicator/IT.NET.USER.ZS) The values are the percentage of the total populations in the region that has access to internet.
```{r include=FALSE}
internetuser <- read_tsv("INTERNETUSER.csv")
internetuser %>% head()
```


============================
Analysis
============================


```{r include=FALSE}
#**COVID CASES IN 2020 AND 2021**
#COVID cases in 2020 and 2021, plotted
covid_cases %>% filter(date > '2020-01-01') %>% filter(date < '2021-12-30') %>% distinct(date, .keep_all = TRUE) %>% ggplot() + geom_point(aes(x=date, y=total_cases))
#Description: uncovering total COVID case trends in 2020-2021. There appears to be a spike in July 2020 and July 2021.
#Total cases vs population density per country, plotted
covid_cases %>% filter(population_density<40 & population_density>10) %>% 
  ggplot() +
  geom_line(aes(population_density,total_cases_per_million))
#Description: analyzing how a greater population density might increase the risk of COVID transmission. Upon inspection, it is clear that no real trend is apparent.
#**EFFECTIVENESS OF VACCINE**
#Death rate vs total vaccinations per hundred, plotted
deathrate_vaccs <- inner_join(deathrate, covid_vaccs, by = c("name"="location")) %>% filter(date=="2021-12-01")
deathrate_vaccs  %>% ggplot(aes( total_vaccinations_per_hundred, value)) + geom_point() + geom_smooth(method = "lm", se = FALSE)  
#Description: analyzing whether or not vaccinations decreased the death rate in regions. Upon inspection, it is clear that no real trend is apparent. Thus, the data shows that the death rate numbers do not decrease as more COVID vaccines are administered. 
#Potential bias: COVID deaths represent a small subset of the total death rates of a region, so that might explain why there did not seem to be a noticeable drop in death rates as the number of vaccines administered increased.
#Cardiovascular death rate vs total vaccinations per hundred, plotted
covid_cases %>% filter(date=="2021-12-01") %>% ggplot(aes(total_vaccinations_per_hundred, cardiovasc_death_rate)) + geom_point() + geom_smooth(method = lm, se = FALSE)
#Finding correlation between cardiovascular death rate and total vaccinations
cor(covid_cases$cardiovasc_death_rate,covid_cases$total_vaccinations_per_hundred,use="complete.obs")
#Description: analyzing whether or not vaccinations decreased cardiovascular-related death rates in regions. By narrowing down death rate data, it is evident that the number of cardiovascular-related death rates dropped off significantly as the number of vaccines administered increased. Coincidentally, COVID has shown to have significant impacts on the cardiovascular system of victims. In addition, a correlation coefficient of -0.37 indicates a negative relationship between the two variables. Thus, we can conclude that the vaccine successfully limits a major symptom of COVID by a significant margin. 
#New cases vs total vaccinations, plotted
covid_cases %>% filter( date == "2021-12-01") %>% ggplot(aes(total_vaccinations, new_cases_per_million)) + geom_point() + geom_smooth(method = lm, se = FALSE)
cor(covid_cases$new_cases_per_million,covid_cases$total_vaccinations,use="complete.obs")
#Description: analyzing if countries with a large amount of vaccinations have a lower rate of new cases per million in late 2021. Upon inspection alone, it is evident that the number of new cases per million are lower in a country with a higher amount of vaccines administered. A correlation coefficient of -0.06 indicates a negative relationship between the two variables. Thus, we can conclude that a higher number of vaccines administered in a country leads to a proportionately lower rate of new cases
#**EXTERNAL - ECONOMIC AND LIFESTYLE - IMPACTS OF COVID**
#Unemployment in 2017 by region, plotted
unemployment_population<-left_join(unemployment,population,by=c("Country"="name"))
unemployment_population %>% filter(`Date of Information` == 	"2017 est.") %>% group_by(region) %>% summarise(unemployment_percentage = mean(`(%)`))%>% ggplot() + geom_point(aes(x=reorder(region, unemployment_percentage), unemployment_percentage)) + coord_flip()
#Description: analyzing whether or not COVID has an impact on the unemployment rates of a region. Regions such as North America had an unemployment rate of ~5% in 2017. In 2020, the North America unemployment rate was ~8%, but fell back down closer to ~5% in 2021. However, other major regions such as Europe and South Asia did not suffer major fluctuations in their unemployment rates due to COVID. Thus, there is not sufficient evidence to suggest a spike in global unemployment rates due to COVID.Reference for 2020 and 2021 data (not provided by unemployment_population dataset): https://www.statista.com/statistics/279790/unemployment-rate-in-seclected-world-regions/
#Urbanization in 2019 vs 2020, plotted
urbanization %>% ggplot() + geom_point(aes(y=`2020`, x=`Country Code`), method = "lm", colour = "Red") +  geom_point(aes(y=`2019`, x=`Country Code`), method = "lm", colour = "Green")
#Description: analyzing urbanization trends before and after COVID. Upon inspection, it is clear that the percentage of the total population - in every country - in urban areas is consistent before and after COVID; there are no sudden fluctuations or dramatic changes cause by the pandemic. 
#Number of internet users in 2020 vs 2019, plotted
internetuser %>% ggplot() + geom_point(aes(y=`2020`, x=`Country Code`), method = "lm", colour = "Red") +  geom_point(aes(y=`2019`, x=`Country Code`), method = "lm", colour = "Green")
#Description: analyzing whether or not COVID had a significant impact on the number of internet users. By plotting the number of internet users - by country - in 2020 (red) vs 2019 (green), it is evident that the number of internet users were much higher in 2020. This can be attributed to the lockdown, which forced individuals to stay inside their homes. This decreased the time they spent outdoors and, in return, increased the amount of time they spent indoors on the internet. In addition, many companies transitioned to a remote work structure where employees could work from home on their computers. This may explain why the unemployment numbers did not show any significant changes from COVID, but the number of internet users skyrocketed.
#Government response vs happiness score, plotted
response_happiness<-left_join(covid_response_2020,happinessscore,by=c("CountryName"="Country name")) %>% select(CountryName,GovernmentResponseIndex,`Ladder score`)
response_happiness %>%  ggplot() + geom_smooth(aes(GovernmentResponseIndex,`Ladder score`),method = "lm")
#Finding correlation between response and happiness
cor(response_happiness$GovernmentResponseIndex,response_happiness$`Ladder score`,use="complete.obs")
#Description: analyzing if overall happiness of the population in a region increases as the government's response to COVID increases. A correlation coefficient of 0.22 indicates is a positive relationship between the two variables. Thus, as the government's response to COVID is higher, the population will - in general - be happier.
#**COMPARISON**
#Comparing infection rates vs. government response between Canada and China
response_graph <- covid_response %>%
  filter(location == "China", date > "2021-09-01") %>%
  select(c(date, response_index)) %>%
  arrange(date)
covid_cases %>%
  filter(location == "China", date > "2021-09-01") %>%
  select(c(date, total_cases)) %>%
  arrange(date) %>%
  ggplot() +
  geom_bar(aes(date, total_cases) +
  geom_line(data = response_graph, aes(x = date, y = response_index)))
response_graph <- covid_response %>%
  filter(location == "Canada", date > "2021-09-01") %>%
  select(c(date, response_index)) %>%
  arrange(date)
covid_cases %>%
  filter(location == "Canada", date > "2021-09-01") %>%
  select(c(date, total_cases)) %>%
  arrange(date) %>%
  ggplot() +
  geom_bar(aes(date, total_cases) +
  geom_line(data = response_graph, aes(x = date, y = response_index)))
response_graph <- covid_response %>%
  filter(location == "China", date < "2021-05-01") %>%
  select(c(date, response_index)) %>%
  arrange(date)
covid_cases %>%
  filter(location == "China", date < "2021-05-01") %>%
  select(c(date, total_cases)) %>%
  arrange(date) %>%
  ggplot() +
  geom_bar(aes(date, total_cases) +
  geom_line(data = response_graph, aes(x = date, y = response_index)))
response_graph <- covid_response %>%
  filter(location == "Canada", date < "2021-05-01") %>%
  select(c(date, response_index)) %>%
  arrange(date)
covid_cases %>%
  filter(location == "Canada", date < "2021-05-01") %>%
  select(c(date, total_cases)) %>%
  arrange(date) %>%
  ggplot() +
  geom_bar(aes(date, total_cases) +
  geom_line(data = response_graph, aes(x = date, y = response_index)))
```
